---
title: "Untitled"
output: html_document
date: "2026-07-12"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}

rm(list = ls())
library(data.table)
library(dplyr)
library(ggplot2)
library(ggh4x)
library(RTMB)
library(LaMa)
library(clue)
library(actigraph.sleepr)
library(purrr)
library(lubridate)
library(nparACT)

```

# Table 1

```{r}

scores <- read.csv("scores.csv")
setDT(scores)
scores[, group := fifelse(grepl("condition", number), "Depression", "Control")]

```


# Actigraph data

```{r}

# conditions (depressed)

# adjust as needed to correct directories
files <- list.files("/Downloads/depresjon/condition/")
path <- "/Downloads/depresjon/condition/"

condition_all <- lapply(files, function(f){
  d <- read.csv(paste0(path, f))
  d$subject <- tools::file_path_sans_ext(f)
  d
})


# control

# adjust as needed
control_files <- list.files("/Downloads/depresjon/control/")
control_path <- "/Downloads/depresjon/control/"

control_all <- lapply(control_files, function(f){
  d <- read.csv(paste0(control_path, f))
  d$subject <- tools::file_path_sans_ext(f)
  d
})

```



# Helper functions

```{r}

# create 15 minute bins for sleep-wake cycle
aggregate_15min <- function(d){
  d %>%
    mutate(bin = floor_date(timestamp, "15 minutes")) %>%
    group_by(bin) %>%
    summarise(
      activity = mean(activity),
      label = if_else(mean(sleep == "S") > 0.5, "S", "W"),
      .groups = "drop"
    )
}

# take into account that after data cleaning, one subject can have two recordings
# so take average for subject weighted by length of recording (for overall fit)
npar_subj <- function(npar, group){
  npar %>%
    group_by(id) %>%
    summarise(across(c(IS, IV, RA), ~ weighted.mean(.x, n_epochs)), .groups = "drop") %>%
    mutate(group = group)
}

act_subj <- function(segs, group){
  do.call(rbind, lapply(segs, function(s)
    data.frame(id = s$id[1], n = nrow(s), activity = mean(s$activity)))) %>%
    group_by(id) %>%
    summarise(activity = weighted.mean(activity, n), .groups = "drop") %>%
    mutate(group = group)
}

bin_segments <- function(segs){
  lapply(segs, function(s){
    b <- aggregate_15min(s)
    b$timestamp <- b$bin
    b$id <- s$id[1]
    b$seg <- s$seg[1]
    b
  })
}


```


```{r}

settings <- data.frame(epochlength = 60L)
min_epochs <- 48 * 60

condition_segments <- unlist(lapply(seq_along(condition_all), function(n){
  d <- condition_all[[n]] %>%
    mutate(timestamp = as.POSIXct(timestamp, tz = "Europe/Oslo"),
           axis1 = activity, axis2 = 0, axis3 = 0)
  
  d <- d[!is.na(d$timestamp), ]
  d <- d[!duplicated(d$timestamp), ]

  # Choi for non-wear
  periods <- apply_choi(tbl_agd(d, settings),
                        use_magnitude = FALSE, min_period_len = 540)

  # Drop non-wear epochs
  if (nrow(periods) > 0){
    nonwear <- Reduce(`|`, Map(function(s, e) d$timestamp >= s & d$timestamp <= e,
                           periods$period_start, periods$period_end))
    d <- d[!nonwear, ]
  }
  if (nrow(d) == 0) return(NULL)

  # Continuous segments of at least 48h
  gap <- c(TRUE, diff(as.numeric(d$timestamp)) != 60)
  segs <- split(d, cumsum(gap))
  segs <- segs[sapply(segs, nrow) >= min_epochs]
  if (length(segs) == 0) return(NULL)

  # Cole-Kripke per segment
  lapply(seq_along(segs), function(i){
    s <- apply_cole_kripke(tbl_agd(segs[[i]], settings))
    s$id <- segs[[i]]$subject[1]
    s$seg <- i
    s
  })
}), recursive = FALSE)

```


```{r}

control_segments <- unlist(lapply(seq_along(control_all), function(n){
  d <- control_all[[n]] %>%
    mutate(timestamp = as.POSIXct(timestamp, tz = "Europe/Oslo"),
           axis1 = activity, axis2 = 0, axis3 = 0)
  
  d <- d[!is.na(d$timestamp), ]
  d <- d[!duplicated(d$timestamp), ]

  # Choi for non-wear
  periods <- apply_choi(tbl_agd(d, settings),
                        use_magnitude = FALSE, min_period_len = 540)

  # Drop non-wear epochs
  if (nrow(periods) > 0){
    nonwear <- Reduce(`|`, Map(function(s, e) d$timestamp >= s & d$timestamp <= e,
                           periods$period_start, periods$period_end))
    d <- d[!nonwear, ]
  }
  if (nrow(d) == 0) return(NULL)

  # Continuous segments of at least 48h
  gap <- c(TRUE, diff(as.numeric(d$timestamp)) != 60)
  segs <- split(d, cumsum(gap))
  segs <- segs[sapply(segs, nrow) >= min_epochs]
  if (length(segs) == 0) return(NULL)

  # Cole-Kripke per segment
  lapply(seq_along(segs), function(i){
    s <- apply_cole_kripke(tbl_agd(segs[[i]], settings))
    s$id <- segs[[i]]$subject[1]
    s$seg <- i
    s
  })
}), recursive = FALSE)

```


```{r}

# condition and control bins will contain the cleaned segments (dropped non-wear, Cole-Kripke labels, split into continuous segments and aggregated to 15 minutes)
# do sensitivity tests on condition and control all

condition_bins <- bin_segments(condition_segments)
control_bins <- bin_segments(control_segments)

```


```{r}

# condition_all and control_all do NOT get aggregated to 15-min epochs
# use for sensitivity checks
# but remove non-wear as this represents invalid data

condition_all <- lapply(condition_all, function(d){
  d <- d %>%
    mutate(timestamp = as.POSIXct(timestamp, tz = "Europe/Oslo"),
           axis1 = activity, axis2 = 0, axis3 = 0)
  d <- d[!is.na(d$timestamp), ]
  d <- d[!duplicated(d$timestamp), ]
  periods <- apply_choi(tbl_agd(d, settings), use_magnitude = FALSE, min_period_len = 540)
  if (nrow(periods) > 0){
    nonwear <- Reduce(`|`, Map(function(s, e) d$timestamp >= s & d$timestamp <= e,
                               periods$period_start, periods$period_end))
    d <- d[!nonwear, ]
  }
  d
})

control_all <- lapply(control_all, function(d){
  d <- d %>%
    mutate(timestamp = as.POSIXct(timestamp, tz = "Europe/Oslo"),
           axis1 = activity, axis2 = 0, axis3 = 0)
  d <- d[!is.na(d$timestamp), ]
  d <- d[!duplicated(d$timestamp), ]
  periods <- apply_choi(tbl_agd(d, settings), use_magnitude = FALSE, min_period_len = 540)
  if (nrow(periods) > 0){
    nonwear <- Reduce(`|`, Map(function(s, e) d$timestamp >= s & d$timestamp <= e,
                               periods$period_start, periods$period_end))
    d <- d[!nonwear, ]
  }
  d
})

```



# Plot


```{r}

control1 <- control_segments[[1]] # example subject

ggplot(control1[1021:2461, ], aes(x = timestamp, y = log1p(activity), fill = sleep)) +
  geom_col() +
  scale_fill_manual(values = c(S = "#3B6EA5", W = "#E8A33D")) +
  labs(x = "Time", y = "Activity count", fill = NULL) +
  theme_minimal()

```

# RA, IS and IV


```{r}

# compute using nparACT
# 15 minute epochs and 24 hour reference (standard)

condition_npar <- do.call(rbind, lapply(condition_bins, function(s){
  assign("dat", data.frame(timestamp = s$timestamp, activity = s$activity),
         envir = .GlobalEnv)
  r <- nparACT_base("dat", SR = 1/900, plot = FALSE)
  data.frame(id = s$id[1], seg = s$seg[1], n_epochs = nrow(s), r)
}))

control_npar <- do.call(rbind, lapply(control_bins, function(s){
  assign("dat", data.frame(timestamp = s$timestamp, activity = s$activity),
         envir = .GlobalEnv)
  r <- nparACT_base("dat", SR = 1/900, plot = FALSE)
  data.frame(id = s$id[1], seg = s$seg[1], n_epochs = nrow(s), r)
}))

```


```{r}

# test difference

npar_all <- bind_rows(npar_subj(condition_npar, "Depression"),
                      npar_subj(control_npar, "Control"))

## test

wilcox_is <- wilcox.test(npar_all$IS ~ npar_all$group, exact = FALSE)
wilcox_iv <- wilcox.test(npar_all$IV ~ npar_all$group, exact = FALSE)
wilcox_ra <- wilcox.test(npar_all$RA ~ npar_all$group, exact = FALSE)

# apply Holm correction below (multiple tests)

setDT(npar_all)
summary <- npar_all[, .(med_IS = median(IS), q1_IS = quantile(IS, .25), q3_IS = quantile(IS, .75),
                        med_IV = median(IV), q1_IV = quantile(IV, .25), q3_IV = quantile(IV, .75),
                        med_RA = median(RA), q1_RA = quantile(RA, .25), q3_RA = quantile(RA, .75)),
                    by = group]
summary

```




## Sensitivity check - no non-wear removal and no aggregation to 15 minute bins


```{r}

npar_raw <- function(all, group){
  do.call(rbind, lapply(all, function(d){
    assign("dat", data.frame(timestamp = as.POSIXct(d$timestamp, tz = "Europe/Oslo"),
                             activity  = d$activity), envir = .GlobalEnv)
    data.frame(id = d$subject[1], group = group,
               nparACT_base("dat", SR = 1/60, plot = FALSE))
  }))
}

npar_noagg <- rbind(npar_raw(condition_all, "Depression"),
                     npar_raw(control_all, "Control"))

```


```{r}

# test difference

wilcox_is_sens <- wilcox.test(npar_noagg$IS ~ npar_noagg$group, exact = FALSE)
wilcox_iv_sens <- wilcox.test(npar_noagg$IV ~ npar_noagg$group, exact = FALSE)
wilcox_ra_sens <- wilcox.test(npar_noagg$RA ~ npar_noagg$group, exact = FALSE)

wilcox_is_sens

wilcox_iv_sens

wilcox_ra_sens

```


# Total activity count

```{r}

prof <- bind_rows(
  bind_rows(condition_bins) %>% mutate(group = "Group with depressive disorders"),
  bind_rows(control_bins) %>% mutate(group = "Control group")
) %>%
  mutate(tod = as.numeric(format(bin, "%H")) + as.numeric(format(bin, "%M")) / 60) %>%
  group_by(group, id, tod) %>%
  summarise(activity = mean(activity), .groups = "drop") %>%
  group_by(group, tod) %>%
  summarise(activity = mean(activity), .groups = "drop")

activity_plot <- ggplot(prof, aes(tod, activity, colour = group, linetype = group)) +
  geom_line(linewidth = 0.8) +
  scale_x_continuous(breaks = seq(0, 24, 4),
                     labels = sprintf("%02d:00", seq(0, 24, 4)),
                     expand = c(0, 1)) +
  scale_colour_manual(values = c("Group with depressive disorders" = "#1B7837", "Control group" = "#333333")) +
  labs(x = "Time of day", y = "Mean activity count / 15-min epoch",
       colour = NULL, linetype = NULL) +
  theme_classic(base_size = 12) +
  theme(legend.position = "top")

ggsave(filename = "activity_plot.pdf", activity_plot, width = 7, height = 5)


```


```{r}

act <- bind_rows(act_subj(condition_bins, "Depression"),
                 act_subj(control_bins, "Control"))

t_act <- t.test(activity ~ group, data = act)

act %>% group_by(group) %>% summarise(mean = mean(activity), sd = sd(activity))

```


# Holm correction

```{r}

p_vals <- c(wilcox_is$p.value, wilcox_iv$p.value, wilcox_ra$p.value, t_act$p.value)

p.adjust(p_vals, method = "holm") # adjusted

p_vals # original


```


## Holm correction for sensitivity checks

```{r}

p_vals_sens <- c(wilcox_is_sens$p.value, wilcox_iv_sens$p.value, wilcox_ra_sens$p.value)

p.adjust(p_vals_sens, method = "holm") # Holm correctio but for sensitivity checks

p_vals_sens # original ones

```


# Hidden Markov Models

## Check what emissions appropriate

```{r}

condition <- lapply(condition_segments, function(s){
  b <- aggregate_15min(s)
  b$id <- s$id[1]
  b$seg <- s$seg[1]
  b
})

control <- lapply(control_segments, function(s){
  b <- aggregate_15min(s)
  b$id <- s$id[1]
  b$seg <- s$seg[1]
  b
})

```


```{r}

emis <- bind_rows(
  bind_rows(condition) %>% mutate(group = "Depression"),
  bind_rows(control) %>% mutate(group = "Control")) 

brk <- c(-0.5, 0.5, seq(10, quantile(emis$activity, 0.99), length = 40))

ggplot(emis, aes(activity, fill = label)) +
  geom_histogram(aes(y = after_stat(density)), breaks = brk,
                 alpha = 0.5, position = "identity", colour = NA) +
  coord_cartesian(xlim = c(0, 800), ylim = c(0, 0.008)) +
  facet_wrap(~ group) +
  scale_fill_manual(values = c(S = "#3B6EA5", W = "#E8A33D"),
                    labels = c(S = "Sleep", W = "Wake")) +
  labs(x = "Mean activity count per 15-min epoch", y = "Density", fill = NULL) +
  theme_classic()

```


## HMM

## HMM for condition


```{r}

source("git_master_act_functions.R")

condition_data <- lapply(condition_bins, function(d){
  matrix(as.numeric(d$activity), ncol = 1)
})

J <- 2
mu0 <- matrix(c(20, 250), nrow = 1) # K x J: rest low, active high (counts)
sigma0 <- matrix(c(30, 200), nrow = 1)    # K x J: SD in counts
z0 <- matrix(c(0.15, 0.001), nrow = 1) # K x J: P(exact zero) per state
tpm0 <- matrix(c(0.9, 0.1,
                   0.1, 0.9), nrow = 2, byrow = TRUE)

hmm_fit <- mle_hmm_multiple(mu0, sigma0, z0, tpm0, J, condition_data)

```


```{r}

fitted_states <- my_viterbi_decoder(hmm_fit$delta, hmm_fit$tpm, condition_data, J, 
                                    hmm_fit$mu, hmm_fit$sigma, hmm_fit$z ,model = "HMM")

cole_condition_labels <- rbindlist(condition_bins)$label

fitted_states <- unlist(fitted_states)

cm <- table(fitted_states, cole_condition_labels)

assign <- solve_LSAP(cm, maximum = TRUE)
aligned_states <- assign[fitted_states]

cm <- table(cole_condition_labels, aligned_states)

cat("State assignment is: \n")
assign

cat("Proportion of time spent in each state for condition: \n")
table(fitted_states)/length(fitted_states)

cat("Activity counts for condition: \n")
hmm_fit$mu

hmm_fit$sigma

hmm_fit$z

cat("t.p.m. for conditions: \n")
hmm_fit$tpm

```


```{r}

n <- sum(cm) 
nc <- nrow(cm)
diag <- diag(cm) 
rowsums <- apply(cm, 1, sum)
colsums <- apply(cm, 2, sum) 
p <- rowsums / n 
q <- colsums / n 

accuracy <- sum(diag) / n 

cat("Agreement with Cole-Kripke for conditions: \n")
accuracy

# accuracy 
# 
# precision = diag/colsums 
# recall = diag/rowsums 
# f1 = 2 * precision * recall / (precision + recall) 
# data.frame(precision, recall, f1) 


```

## HMM for control


```{r}

control_data <- lapply(control_bins, function(d){
  matrix(as.numeric(d$activity), ncol = 1)
})

# same as for condition

# J <- 2
# mu0 <- matrix(c(20, 250), nrow = 1) # K x J: rest low, active high (counts)
# sigma0 <- matrix(c(30, 200), nrow = 1)    # K x J: SD in counts
# z0  <- matrix(c(0.15, 0.001), nrow = 1) # K x J: P(exact zero) per state
# tpm0 <- matrix(c(0.9, 0.1,
#                    0.1, 0.9), nrow = 2, byrow = TRUE)

hmm_fit_control <- mle_hmm_multiple(mu0, sigma0, z0, tpm0, J, control_data)

```


```{r}

fitted_states <- my_viterbi_decoder(hmm_fit_control$delta, hmm_fit_control$tpm, control_data, J, 
                                    hmm_fit_control$mu, hmm_fit_control$sigma, hmm_fit_control$z ,model = "HMM")

cole_control_labels <- rbindlist(control_bins)$label

fitted_states <- unlist(fitted_states)

cm <- table(fitted_states, cole_control_labels)

assign <- solve_LSAP(cm, maximum = TRUE)
aligned_states <- assign[fitted_states]

cm <- table(cole_control_labels, aligned_states)

cat("State assignment is: \n")
assign

cat("Proportion of time spent in each state for control: \n")
table(fitted_states)/length(fitted_states)
  

cat("Activity counts for controls: \n")
hmm_fit_control$mu

hmm_fit_control$sigma

hmm_fit_control$z

cat("t.p.m. for controls: \n")
hmm_fit_control$tpm

```


```{r}

n <- sum(cm) 
nc <- nrow(cm)
diag <- diag(cm) 
rowsums <- apply(cm, 1, sum) 
colsums <- apply(cm, 2, sum) 
p <- rowsums / n 
q <- colsums / n 

accuracy <- sum(diag) / n 

cat("Agreement with Cole-Kripke for controls: \n")
accuracy

# accuracy 
# 
# precision = diag/colsums 
# recall = diag/rowsums 
# f1 = 2 * precision * recall / (precision + recall) 
# data.frame(precision, recall, f1) 

```


## nb-HSMM

## nb-HSMM for condition


```{r}

R_vec <- c(40, 80)

nb_size0 <- c(2, 2)       
nb_mu0 <- c(10, 10) # match HMM init t.p.m.

cond_tpm0 <- matrix(c(0, 1,
                      1, 0), nrow = 2, byrow = TRUE) # since J = 2

nbhsmm_fit <- mle_nbhsmm_multiple(mu0, sigma0, z0, nb_size0, nb_mu0, cond_tpm0,
                                  J, condition_data, R_vec)

```

```{r}

fitted_states <- my_viterbi_decoder(nbhsmm_fit$delta, nbhsmm_fit$tpm, condition_data, J, 
                                    nbhsmm_fit$mu, nbhsmm_fit$sigma, nbhsmm_fit$z ,model = "HSMM", R_vec = R_vec)


fitted_states <- unlist(fitted_states)

cm <- table(fitted_states, cole_condition_labels)

assign <- solve_LSAP(cm, maximum = TRUE)
aligned_states <- assign[fitted_states]

cm <- table(cole_condition_labels, aligned_states)

cat("State assignment is: \n")
assign

cat("Proportion of time spent in each state for condition: \n")
table(fitted_states)/length(fitted_states)

cat("Activity counts for condition: \n")
nbhsmm_fit$mu

nbhsmm_fit$sigma


nbhsmm_fit$z



cat("NB distributions: \n")
nbhsmm_fit$nb_mu



n <- sum(cm) # number of instances
diag <- diag(cm) # number of correctly classified instances per class 
accuracy <- sum(diag) / n 

cat("Agreement with Cole-Kripke for conditions: \n")
accuracy

```


## nb-HSMM for control


```{r}


nbhsmm_fit_control <- mle_nbhsmm_multiple(mu0, sigma0, z0, nb_size0, nb_mu0, cond_tpm0,
                                  J, control_data, R_vec)


```

```{r}

fitted_states <- my_viterbi_decoder(nbhsmm_fit_control$delta, nbhsmm_fit_control$tpm, control_data, J, 
                                    nbhsmm_fit_control$mu, nbhsmm_fit_control$sigma, nbhsmm_fit_control$z ,model = "HSMM", R_vec = R_vec)


fitted_states <- unlist(fitted_states)

cm <- table(fitted_states, cole_control_labels)

assign <- solve_LSAP(cm, maximum = TRUE)
aligned_states <- assign[fitted_states]

cm <- table(cole_control_labels, aligned_states)

cat("State assignment is: \n")
assign

cat("Proportion of time spent in each state for condition: \n")
table(fitted_states)/length(fitted_states)

cat("Activity counts for control: \n")
nbhsmm_fit_control$mu


nbhsmm_fit_control$sigma


nbhsmm_fit_control$z

cat("NB distributions: \n")
nbhsmm_fit_control$nb_mu

n <- sum(cm) # number of instances
diag <- diag(cm) # number of correctly classified instances per class 
accuracy <- sum(diag) / n 

cat("Agreement with Cole-Kripke for control: \n")
accuracy

```



## QREML

## QREML for condition


```{r}

m <- 2
S <- lapply(R_vec, function(Ri) crossprod(makeD(Ri, m)))

# same init as HMM and nb-HSMM
p_list <- lapply(seq_len(J), function(i){
  p <- dgeom(0:R_vec[i], prob = 0.1)
  p / sum(p)
})


par <- natural_to_unc_nonp_hsmm(mu0, sigma0, z0, p_list, cond_tpm0, J)

dat <- list(data_list = condition_data,
            J = J,
            R_vec = R_vec,
            S = S,
            lambda = rep(1e2, J))


qreml_fit <- qreml_capped(pen_nll, par, dat, random = paste0("unc_p", seq_len(J)))
est <- extract_relevant_pars(qreml_fit, model = "nonp", R_vec = R_vec, J = J)

```


```{r}

paths <- my_viterbi_decoder(qreml_fit$delta, est$hmm_tpm, condition_data, J,
                            qreml_fit$mu, qreml_fit$sigma, qreml_fit$z ,model = "HSMM", R_vec = R_vec)

fitted_states <- unlist(paths)

cm <- table(fitted_states, cole_condition_labels)

assign <- solve_LSAP(cm, maximum = TRUE)
aligned_states <- assign[fitted_states]

cm <- table(cole_condition_labels, aligned_states)

```


```{r}

cat("State assignment is: \n")
assign

cat("Proportion of time spent in each state for condition: \n")
table(fitted_states)/length(fitted_states)

cat("Activity counts for condition: \n")
qreml_fit$mu

qreml_fit$sigma

qreml_fit$z

n <- sum(cm) 
diag <- diag(cm) 
accuracy <- sum(diag) / n 

cat("Agreement with Cole-Kripke for condition: \n")
accuracy

```



## QREML for control


```{r}

dat <- list(data_list = control_data,
            J = J,
            R_vec = R_vec,
            S = S,
            lambda = rep(1e2, J))


qreml_fit_control <- qreml_capped(pen_nll, par, dat, random = paste0("unc_p", seq_len(J)))
est_control <- extract_relevant_pars(qreml_fit_control, model = "nonp", R_vec = R_vec, J = J)

```


```{r}

paths <- my_viterbi_decoder(qreml_fit_control$delta, est_control$hmm_tpm, control_data, J,
                            qreml_fit_control$mu, qreml_fit_control$sigma, qreml_fit_control$z ,model = "HSMM", R_vec = R_vec)

fitted_states <- unlist(paths)

cm <- table(fitted_states, cole_control_labels)

assign <- solve_LSAP(cm, maximum = TRUE)
aligned_states <- assign[fitted_states]

cm <- table(cole_control_labels, aligned_states)

```


```{r}

cat("State assignment is: \n")
assign

cat("Proportion of time spent in each state for condition: \n")
table(fitted_states)/length(fitted_states)

cat("Activity counts for condition: \n")
qreml_fit_control$mu
qreml_fit_control$sigma
qreml_fit_control$z

n <- sum(cm) # number of instances
diag <- diag(cm) # number of correctly classified instances per class 
accuracy <- sum(diag) / n 

cat("Agreement with Cole-Kripke for control: \n")
accuracy

```


## Save fitted models for later checks and plotting

```{r}

saveRDS(
  list(hmm_fit = hmm_fit, hmm_fit_control = hmm_fit_control,
       nbhsmm_fit = nbhsmm_fit, nbhsmm_fit_control = nbhsmm_fit_control,
       qreml_fit = qreml_fit, qreml_fit_control = qreml_fit_control,
       est = est, est_control = est_control, # transformed qREML params
       #condition_data = condition_data, control_data = control_data,
       S = S, J = J, R_vec = R_vec),
  "actigraph_fits.rds")

```

# Plot dwell-time


```{r}


R_eval <- 90   
J <- 2
R_vec  <- c(40, 80)   

fits <- list(
  list(obj = hmm_fit, model = "HMM", grp = "Group with depressive disorders", tag = "hmm"),
  list(obj = hmm_fit_control, model = "HMM",grp = "Control group", tag = "hmm"),
  list(obj = nbhsmm_fit, model = "nb-HSMM", grp = "Group with depressive disorders", tag = "nb"),
  list(obj = nbhsmm_fit_control,  model = "nb-HSMM", grp = "Control group", tag = "nb"))

state_lab <- c("Rest", "Active")

dwell_long <- rbindlist(lapply(fits, function(f){
  pars <- extract_relevant_pars(f$obj, model = f$tag, R_vec = R_vec, J = J)
  D    <- build_dwell_matrix(pars, model = f$tag, J = J, R_vec = R_vec, R_eval = R_eval)
  rbindlist(lapply(1:J, function(s){
    data.table(model = f$model, group = f$grp, state = state_lab[s],
               duration = 1:R_eval, prob = D[s, ])
  }))
}))

# lock facet order
dwell_long[, model := factor(model, levels = c("HMM", "nb-HSMM", "qREML-HSMM"))]
dwell_long[, state := factor(state, levels = c("Rest", "Active"))]

hmm_nb_dwell_dep <- ggplot(dwell_long, aes(duration, prob, colour = group, linetype = group)) +
  geom_line(linewidth = 0.7) +
  facet_grid(model ~ state, scales = "free") +
  facetted_pos_scales(
    x = list(
      state == "Rest"   ~ scale_x_continuous(limits = c(0, 50),
                                             breaks = seq(0, 48, 8),
                                             labels = function(x) x / 4),
      state == "Active" ~ scale_x_continuous(limits = c(0, 85),
                                             breaks = seq(0, 80, 16),
                                             labels = function(x) x / 4)
    )
  ) +
  scale_colour_manual(values = c("Group with depressive disorders" = "#1B7837",
                               "Control group" = "#333333"))  +
  labs(x = "Dwell duration (hours)", y = "Estimated probability",
       colour = NULL, linetype = NULL) +
  theme_bw(base_size = 12) +
  theme(legend.position = "top",
        panel.grid.minor = element_blank(),
        strip.text = element_text(face = "plain"))

ggsave(filename = "hmm_nb_dwell_dep.pdf", hmm_nb_dwell_dep, width = 8, height = 6)

```


```{r}

library(ggh4x)

R_eval <- 90
J <- 2
R_vec  <- c(40, 80)

qreml_fits <- list(
  list(obj = qreml_fit, grp = "Group with depressive disorders"),
  list(obj = qreml_fit_control, grp = "Control group")
)

state_lab <- c("Rest", "Active")

qreml_long <- rbindlist(lapply(qreml_fits, function(f){
  pars <- extract_relevant_pars(f$obj, model = "nonp", R_vec = R_vec, J = J)
  D    <- build_dwell_matrix(pars, model = "nonp", J = J, R_vec = R_vec, R_eval = R_eval)
  rbindlist(lapply(1:J, function(s){
    data.table(group = f$grp, state = state_lab[s],
               duration = 1:R_eval, prob = D[s, ])
  }))
}))

qreml_long[, state := factor(state, levels = c("Rest", "Active"))]

qreml_dwell_dep <- ggplot(qreml_long, aes(duration, prob, colour = group, linetype = group)) +
  geom_line(linewidth = 0.9) +
  facet_wrap(~ state, scales = "free") +
  facetted_pos_scales(
    x = list(
      state == "Rest"   ~ scale_x_continuous(limits = c(0, 45),
                                             breaks = seq(0, 48, 8),
                                             labels = function(x) x / 4),
      state == "Active" ~ scale_x_continuous(limits = c(0, 85),
                                             breaks = seq(0, 84, 16),
                                             labels = function(x) x / 4)
    ),
    y = list(
      state == "Rest"   ~ scale_y_continuous(limits = c(0, 0.13)),
      state == "Active" ~ scale_y_continuous(limits = c(0, 0.25))
    )
  ) +
  scale_colour_manual(values = c("Group with depressive disorders" = "#1B7837", "Control group" = "#333333")) +
  labs(x = "Dwell-time (hours)", y = "Estimated probability",
       colour = NULL, linetype = NULL) +
  theme_bw(base_size = 13) +
  theme(legend.position = "top",
        panel.grid.minor = element_blank(),
        strip.text = element_text(face = "plain"))


ggsave(filename = "qreml_dwell_dep.pdf", qreml_dwell_dep, width = 8, height = 6)

```


## Example illustration on two single subjects


```{r}


# rebuild per-subject decoded lists
fitted_states <- my_viterbi_decoder(qreml_fit$delta, est$hmm_tpm, condition_data, J,
                                    qreml_fit$mu, qreml_fit$sigma, qreml_fit$z,
                                    model = "HSMM", R_vec = R_vec)
fitted_states_control <- my_viterbi_decoder(qreml_fit_control$delta, est_control$hmm_tpm,
                                            control_data, J,
                                            qreml_fit_control$mu, qreml_fit_control$sigma,
                                            qreml_fit_control$z, model = "HSMM", R_vec = R_vec)

dep_idx <- 8L
con_idx <- 8L
state_lab <- c("Rest", "Active")  

build_panel <- function(bins, decoded, idx, group_lab){
  b <- as.data.table(bins[[idx]])
  s <- decoded[[idx]]
  n <- min(nrow(b), length(s))


  start <- which(as.numeric(format(b$bin, "%H")) +
                 as.numeric(format(b$bin, "%M")) / 60 >= 9)[1]
  w <- start:(start + 96 - 1)      
  w <- w[w <= n]                   

  data.table(
    group    = group_lab,
    elapsed  = (seq_along(w) - 1) * 0.25,   # hours from 09:00
    activity = as.numeric(b$activity[w]),
    state    = factor(state_lab[s[w]], levels = c("Rest", "Active"))
  )
}

panel <- rbind(
  build_panel(condition_bins, fitted_states, dep_idx, "Subject with a depressive disorder"),
  build_panel(control_bins,   fitted_states_control, con_idx, "Control subject")
)
panel[, group := factor(group, levels = c("Subject with a depressive disorder", "Control subject"))]


panel[, run := rleid(state), by = group]
segs <- panel[, .(xstart = min(elapsed), state = state[1]), by = .(group, run)]
segs[, xend := shift(xstart, type = "lead"), by = group]
grp_max <- panel[, .(tmax = max(elapsed)), by = group]
segs[grp_max, on = "group", tmax := i.tmax]
segs[is.na(xend), xend := tmax]
segs[, tmax := NULL]


brk <- seq(0, 24, 3)
lab <- sprintf("%02d:00", (9 + brk) %% 24)

act_decoded <- ggplot() +
  geom_rect(data = segs,
            aes(xmin = xstart, xmax = xend, ymin = -Inf, ymax = Inf, fill = state)) +
  geom_line(data = panel, aes(elapsed, activity), linewidth = 0.6, colour = "grey15") +
  facet_wrap(~ group, ncol = 1) +
  scale_fill_manual(values = c(Rest = "#D5E3F2", Active = "#FCE4CF"), name = NULL)  +
  scale_x_continuous(breaks = brk, labels = lab, expand = c(0, 0)) +
  labs(x = "Time of day", y = "Activity count") +
  theme_bw(base_size = 12) +
  theme(legend.position = "top",
        panel.grid.minor = element_blank(),
        strip.text = element_text(face = "plain"))

ggsave(filename = "act_decoded.pdf", act_decoded, width = 10, height = 6)

```




